Refinements for open automata

### 2023-04-28 2021-Refinement/paper/refinementarticleV2.tex
qques corrections

The OA  modelling the behaviour of such a system  using an  unbounded circular/ring buffer is depicted in Fig. \ref{Fig:RefineOA}.  The automaton has a single state with  two holes: $\texttt{P}$ and  $\texttt{Q}$ that are the two interacting processes.  $\texttt{l}$   indicates the next available position for  enqueuing an element  and  $\texttt{f}$ is the position that contains the next element to be dequeued. %\footnote{The  arrow  without any state of origin is a graphical convention to indicate that the pointers are initialized at the beginning}. 
The buffer reacts to a push from $\texttt{P}$ and enqueues it. Similarly, whenever   $\texttt{Q}$ pops an element, it dequeues it. Additionally, whenever $\texttt{Q}$ produces an item, it is exposed as an external observable  $\texttt{print}$ action. 
When any process do its internal computation, it is exposed externally as unobservable action $\tau$. 
\begin{figure}[!tb]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/FIFORing.pdf}
   \caption{Open Automaton for the \texttt{prod-cons} system using FIFO circular buffer.
    \label{Fig:RefineOA}} 
\end{figure}
\qed
\end{example}

@@

OA are partially specified automata, that partiality comes  from the holes.
A hole can be seen as a port in which we can plug an OA.
The plugging operation is called composition.
The composition of OA was already implicitly defined by the means of composition on pNets in previous work \cite{henrio:01299562} but never  formalised on OA.\footnote{The definition of composition below is a direct translation of what happens with pNets composition \cite{henrio:01299562} without  introducing pNets.}

@@

\begin{example} \label{Example:prodandcons} Fig. \ref{Fig:procandcons} shows a producer automaton  and a consumer  automaton that can be used  to fill the holes $\texttt{P}$ and $\texttt{Q}$ of \texttt{prod-cons} defined in Example~\ref{ex:fifo-system}.
 \begin{figure}[!ht]
 \centering
   \includegraphics[width=.35\textwidth]{Figures/P-proc.pdf}\hfill 
   \includegraphics[width=.45\textwidth]{Figures/Q-proc.pdf}
   \caption{(Left) A \texttt{producer}.  It produces one item at a time and pushes it.  (Right)  A \texttt{consumer}. It pops an item, does some work and pushes the result. \label{Fig:procandcons}}
\end{figure}

The OA on Fig. \ref{Fig:ComposeOA} is the composition of the system in Fig.~\ref{Fig:RefineOA} and the producer in Fig. \ref{Fig:procandcons}. The composition consists of two states (the product of the states of both automata). The transitions from one state to another come from  the synchronisation  of the transitions of the encompassing automaton with those of the producer filling the hole $\texttt{P}$. Transitions involving \texttt{Q} remain unchanged. \qed

\begin{figure}[!ht]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/compos-FIFO-Pro.pdf}
   \caption{Open Automaton for \texttt{prod-cons}[\texttt{P}/\texttt{producer}]. 
   \label{Fig:ComposeOA}} 
\end{figure}
\end{example}

\subsection{Relations between open automata}
A relation (bisimulation, refinement, etc.) between OA requires to compare their states. To do so we will suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on open automata configurations] Suppose $V_1$ and $V_2$ are disjoint.
A relation on configurations of two OA \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
%\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the state of the automaton, i.e. if the variables of the OA verify a certain formula. 
In other words, for each pair of states, a boolean formula that may refer to the  variables of each of the two OA states whether the two states are related or not.
Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\).

@@

  \draw (1) edge[align=center] node[below]{\OT{}{}{\alpha(x)}{\{h\mapsto\beta(x)\}}{[x<0]}{(y\leftarrow -x)}}  (2);
  \draw (1) edge[align=center] node[above]{\OT{}{}{\alpha(x)}{\{h\mapsto\beta(x)\}}{[x\geq 0]}{(y\leftarrow x)}} (3);
\end{tikzpicture}
\hfill
\begin{tikzpicture}
  \node[state,inner sep=5pt,minimum size=10pt] (1) {$s_2$};
  \node[state,inner sep=5pt,minimum size=10pt,  right of=1,right=90pt] (2) {$t_2$};
\draw (1) edge[align=center] node[below]{\OT{}{}{\alpha(x)}{\{h\mapsto\beta(x)\}}{True}{(z\leftarrow x)}}  (2);
%  \draw (1) edge[align=center] node[above]{\OT{}{}{\alpha(x)}{h\mapsto\beta(x)}{x\geq 0}{y\leftarrow x}} (3);
\end{tikzpicture}~

@@

Similarly to FH-bisimulation~\cite{fhbisim} we are interested  in finding relations between configurations of two OA that contain variables and holes. 
We rely on a classical notion of simulation inspired from~\cite{fhbisim}. The idea is that two configurations related by a relation; if one state can do a transition, then the other can do a transition too. 
Like for bisimulation, refinement relation characterizes when two states are related, and this  characterisation is expressed as a predicate on the variables of the two automata.
Refinement defines conditions on a predicate $R$ such that $R\mpar{s_1, s_2}$ is a predicate (possibly involving variables of the automata) that is true when the state $s_1$ of $A_1$ refines the state $s_2$ of $A_2$.

@@

\begin{example} \label{example:fifo-system-one}
To illustrate the refinement simulation of OA, we consider a variation on the \texttt{prod-cons} example. Namely,  we suppose that  the two processes  \texttt{P} and  \texttt{Q} communicate through a one-place  buffer. The  corresponding behaviour of this simpler version of the system is specified by the    OA  depicted in Fig. \ref{Fig:SpecOA}, referred to as \texttt{Simprod-cons}. 
\begin{figure}[!t]
 \centering
   \includegraphics[width=.90\textwidth]{Figures/FIFOpen2.pdf}
   \caption{  The \texttt{Simprod-cons} Open Automaton: \texttt{prod-cons} system using one-place  buffer.  \label{Fig:SpecOA}}
\end{figure}
We can easily check that this automaton simulates the one on Fig.~\ref{Fig:RefineOA}.
The simulation relation is $R=\{(r_0,s_0) \mapsto \texttt{l}=\texttt{f}, (r_0,s_1) \mapsto \texttt{f}=\texttt{l}+1\% \texttt{N} \}$ is a simulation. \qed
\end{example}

@@

%\begin{definition}[Non-blocking composition] \label{Def:Non-block}
%Let \(A_i = \OAg[i]\) with \( 0 \leq i \leq n\) be a family of open automata.
%Let \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) and $A= \OAg$.
%
%The composition \(A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) is non-blocking if   \(A \) has a reachability predicate such that, for each reachable configuration, if there is a possible transition in \(A_0\) then there is a possible transition in \(A\):
%\begin{multline*}
% \forall s = \mpar{s_0, s_1,..,s_n} \in S, V \uplus \biguplus_\subbox{t_0 \in \fOT{s_0}} \fvars{t_0} \vdash \\ \left(\reach{A}\mpar{s} \wedge \bigvee_\subbox{t_0 \in \fOT{s_0}} \fguard{t_0} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \right)
%\end{multline*}
%\end{definition}
\begin{definition}[Non-blocking composition] \label{Def:Non-block}
Let \(A_1 \defobject \OAg[1]\), \( A_2 \defobject \OAg[1]\) be two OA.
Let \(A = A_1\subst{A_2}{k}\) and $A= \OAg$.

The composition \(A_1\subst{A_2}{k}\) is non-blocking if   \(A \) has a reachability predicate such that, for each reachable configuration, if there is a possible transition in \(A_1\) then there is a possible transition in \(A\):
\begin{multline*}
 \forall s = \mpar{s_1, s_2} \in S, V \uplus \biguplus_\subbox{t \in \fOT{s_1}} \fvars{t} \vdash \\ \left(\reach{A}\mpar{s} \wedge \bigvee_\subbox{t \in \fOT{s_1}} \fguard{t} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \right)
\end{multline*}
\end{definition}
Like in the definition of refinement (Definition~\ref{def:HoleEqualSim}) we use guards to ensure that the transition can occur. 
In general, one would not want to only consider non-blocking composition as it may reveal a bit restrictive, but it is the best necessary condition that can be identified for compositionality of refinement. It will be used to prove composition theorems in  Section~\ref{sec:prop}. 
Proving refinement relation in absence of non-blocking composition can always be done specifically for the composed automaton under study.
%It is not sufficient to ensure the existence of equivalent transitions in general, but it will be sufficient in the context of refinement.

@@

\begin{property}[Relating simulations]\label{lem-sim-sim}
Hole-equal simulation is a particular case of hole-tracking refinement when  $J_1=J_2=H$.
\end{property}
\TODO{potential ex 4: check if the example with a hole filled is a simulation of the example with holes? at least state \texttt{prod-cons} $\leq_{\{\texttt{P,Q}\}}$ \texttt{simprod-cons}}

@@

\begin{theorem}[Context refinement]\label{thm:ContextRefinement}
Let $A_1$, $A_2$ and $A_3$ be three OA with $\wrel{A_1}{A_2}{H}$. 
Let $J_3$ be the set of holes of $A_3$ and suppose that $k \in H$.
Suppose that \(k \in H\) and that \(A_1\subst{A_3}{k}\) is non-blocking.
We have: \[\wrel{A_1\subst{A_3}{k}}{A_2\subst{A_3}{k}}{J_3 \uplus H \setminus \mbrc{k}}\]
\end{theorem}

Pictorially, the theorem states the following result:

@@

The bisimulation relation that we consider is the one that makes $A_1$ and $A_2$ bisimilar, complemented with identity of configurations for $A_3$.
The proof relies on the fact that, by construction, all transitions of the composed automaton $A_1\subst{A_3}{k}$ are specified by open transitions of $A_1$. For the transitions that do not involve hole $k$, the transition of $A_1\subst{A_3}{k}$ is the same and simulation between $A_1$ and $A_2$ allows us to conclude directly. If the hole $k$ is involved the considered relation  implies that valuations in $A_3$ are equal (i.e., the value for each variable are the same in both valuations), after a transition we should obtain ``equal'' valuations because post-conditions are deterministic.

\begin{example}  Consider again the \texttt{prod-cons} and \texttt{Simprod-cons}  automata given in the examples above.  Since we have \texttt{prod-cons} $\leq_{\{\texttt{P,Q}\}}$ \texttt{simprod-cons},  hence according to Theorem \ref{thm:ContextRefinement}, we get 
\texttt{prod-cons[P/producer]} $\leq_{\{\texttt{Q}\}}$  \texttt{Simprod-cons[P/producer]}. 
The behaviour of \texttt{prod-cons[P/producer]} is shown Fig.~\ref{Fig:ComposeOA}.
The automaton resulting  from the composition of  \texttt{Simprod} and \texttt{producer} is bigger and not shown here. \qed
\end{example}

\begin{theorem}[Congruence]\label{thm:CongurenceRefinement}
Let $A_1$, $A_2$ and $A_3$ be three OA with $\wrel{A_2}{A_3}{H}$. 
  Let $J_3$ be the set of holes of $A_3$ and suppose that \(k \in H\). Suppose additionally that the composition $A_1\subst{A_2}{k}$ is non-blocking.
We have: \[\wrel{A_1\subst{A_2}{k}}{A_1\subst{A_3}{k}}{J_1 \uplus H \setminus \mbrc{k}}\]
\end{theorem}

Consequently, as the simulation is transitive \TODO{TODO Thm 4}

\begin{example}
As an example of the use of this theorem, if we design a refined version of the producer process of Example~\ref{Example:prodandcons} and call it $\texttt{producer}^{*}$. According to  Theorem \ref{thm:CongurenceRefinement}, we have 
\texttt{prod-cons[P/producer]} $\leq_{\{\texttt{Q}\}}$  \texttt{prod-cons[P/$\texttt{producer}^{*}$]}; as well as \texttt{prod-cons[P/producer]} $\leq_{\{\texttt{Q}\}}$  \texttt{Simprod-cons[P/$\texttt{producer}^{*}$]}.

\end{example}

@@

The simulation  ensures the preservation of   
safety properties as deadlock-freeness  and, more generally, all 
all linear temporal logic properties.  The issue is about the properties of simulation, in particular closure properties such as preservation under composition. Some works   have addressed the problem of non-introduction of blocking by composition. For example, in \cite{DIHEGO2020110598} it is presented a refinement relation on process algebra based on failure semantics that is basically different from the (bi)simulation\cite{10.5555/640428.640430}.

@@

\section{Proof of Transitivity for Refinement}\label{sec:proof-transitivity}
\begin{quote}
If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
\end{quote}
\proof 
If $\wrel{A_1}{A_2}{H}$ then there is $R_{12}$ a relation between states
of $A_1$ and of $A_2$;  If $\wrel{A_2}{A_3}{H'}$ then there is $R_{23}$ a relation between states of $A_2$ and of $A_3$. We build a relation between
 states of $A_1$ and of $A_3$ as follows:  for each pair of states $s_1$, $s_3$, for each state $s_2$ such that $R_{12}$ relates $s_1$ and $s_2$, and $R_{23}$ relates $s_2$ and $s_3$.
Let $R_{13}$ be the relation:\\
  \[R_{13}(s_1,s_3)=\bigvee_{s_2\in S_2}\mpar{R_{12}(s_1,s_2) \land R_{23}(s_2,s_3) } \]

@@

This is valid for all $s_{2p} \in (s_{2p})^{p\in P}$, then we have:
\begin{multline*}
 V_1 \uplus V_2 \uplus V_3  \vdash \underbrace{\bigvee_{{p\in P}}\mpar{R_{12}\mpar{s_1, s_{2p}} \wedge R_{23}\mpar{s_{2p},s_3}}}_{R_{13}(s_1,s_3)} \implies  ( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} )
 \lor¢ \lnot( \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} )
\end{multline*}
By removing the $A_2$ variables that have not effect on $A_2$ et $A_3$,
we get the desired result:
\begin{multline*}
 V_1 \uplus  V_3  \vdash R_{13}(s_1,s_3) \implies  ( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} )
 \lor¢ \lnot( \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} )
\end{multline*}
\qed
\end{enumerate}
\section{Proof of Context Refinement (Theorem~\ref{thm:ContextRefinement})}
\begin{quote}
Suppose that $\wrel{A_1}{A_2}{H}$, \(k \in H\) and that \(A_1\subst{A_3}{k}\) is non-blocking. We have: 
\[\wrel{A_1\subst{A_3}{k}}{A_2\subst{A_3}{k}}{J_3 \uplus H \setminus \mbrc{k}}\]
\end{quote}
\smallskip

@@

\section{Proof of Congruence}
Suppose $\wrel{A_2}{A_3}{H}$ and  $k \in H$.  and the composition $A_1\subst{A_2}{k}$ is non-blocking. We have: \[\wrel{A_1\subst{A_2}{k}}{A_1\subst{A_3}{k}}{J_1 \uplus H \setminus \mbrc{k}}\]
\proof
Let us denote by $A_{12}$ (resp. $A_{13}$) the OA resulting from $A_1\subst{A_2}{k}$ (resp. $A_1\subst{A_3}{k}$),  to prove the theorem it is sufficient to prove that there exists a relation between states of the two OA that satisfies the conditions of the Definition \ref{Def:OA-Refinement}.

>>>>

The OA  modelling the behaviour of such a system  using an  unbounded circular/ring buffer is depicted in Fig. \ref{Fig:RefineOA}.  The automaton has a single state with  two holes: $\texttt{P}$ and  $\texttt{Q}$ that are the two interacting processes.  $\texttt{l}$   indicates the next available position for  enqueuing an element  and  $\texttt{f}$ is the position that contains the next element to be dequeued. %\footnote{The  arrow  without any state of origin is a graphical convention to indicate that the pointers are initialized at the beginning}. 
The buffer reacts to a push from $\texttt{P}$ and enqueues it. Similarly, whenever   $\texttt{Q}$ pops an element, it dequeues it. Additionally, whenever $\texttt{Q}$ produces an item, it is exposed as an external observable  $\texttt{print}$ action. 
When any process do its internal computation, it is exposed externally as unobservable action $\tau$. 
\begin{figure}[!tb]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/FIFORing.pdf}
   \caption{Open automaton for the \texttt{prod-cons} system using FIFO circular buffer.
    \label{Fig:RefineOA}} 
\end{figure}
\qed
\end{example}

@@

OA are partially specified automata, the partiality arises  from the holes.
A hole can be seen as a port in which we can plug an OA.
The plugging operation is called composition.
The composition of OA was already implicitly defined by the means of composition on pNets in previous work \cite{henrio:01299562} but never  formalised on OA.\footnote{The definition of composition below is a direct translation of what happens with pNets composition \cite{henrio:01299562} without  introducing pNets.}

@@

\begin{example} \label{Example:prodandcons} Fig. \ref{Fig:procandcons} shows a producer automaton  and a consumer  automaton that can be used  to fill the holes $\texttt{P}$ and $\texttt{Q}$ of \texttt{prod-cons} defined in Example~\ref{Example:fifo-system}.
 \begin{figure}[!ht]
 \centering
   \includegraphics[width=.35\textwidth]{Figures/P-proc.pdf}\hfill 
   \includegraphics[width=.45\textwidth]{Figures/Q-proc.pdf}
   \caption{(Left) A \texttt{producer}.  It produces one item at a time and pushes it.  (Right)  A \texttt{consumer}. It pops an item, does some work and pushes the result. \label{Fig:procandcons}}
\end{figure}

The OA on Fig. \ref{Fig:ComposeOA} is the composition of the system in Fig.~\ref{Fig:RefineOA} and the producer in Fig. \ref{Fig:procandcons} (left). The composition consists of two states (the product of the states of both automata). The transitions from one state to another come from  the synchronisation  of the transitions of the encompassing automaton with those of the producer filling the hole $\texttt{P}$. Transitions involving \texttt{Q} remain unchanged. \qed

\begin{figure}[!ht]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/compos-FIFO-Pro.pdf}
   \caption{Open automaton for \texttt{prod-cons}[\texttt{P}/\texttt{producer}]. 
   \label{Fig:ComposeOA}} 
\end{figure}
\end{example}

\subsection{Relations between open automata}
Establishing semantic equivalences and refinement relations between different OA requires  to compare their states. For this purpose, we suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on open automata configurations] Suppose $V_1$ and $V_2$ are disjoint.
A relation on configurations of two OA \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
%\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the satisfiability of the expression relying their variables i.e., if the variables of the OA verify a certain formula. 
In other words, for each pair of states is attached a boolean formula that may refer to the  variables of each of the two OA states, stating whether the two states are related or not.
Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\).

@@

  \draw (1) edge[align=center] node[below]{\OT{}{}{\alpha(x)}{\{h\mapsto\beta(x)\}}{x<0}{(y\leftarrow -x)}}  (2);
  \draw (1) edge[align=center] node[above]{\OT{}{}{\alpha(x)}{\{h\mapsto\beta(x)\}}{x\geq 0}{(y\leftarrow x)}} (3);
\end{tikzpicture}
\hfill
\begin{tikzpicture}
  \node[state,inner sep=5pt,minimum size=10pt] (1) {$s_2$};
  \node[state,inner sep=5pt,minimum size=10pt,  right of=1,right=90pt] (2) {$t_2$};
\draw (1) edge[align=center] node[below]{\OT{}{}{\alpha(x)}{\{h\mapsto\beta(x)\}}{True}{(z\leftarrow x)}}  (2);
%  \draw (1) edge[align=center] node[above]{\OT{}{}{\alpha(x)}{h\mapsto\beta(x)}{x\geq 0}{y\leftarrow x}} (3);
\end{tikzpicture}~

@@

Similarly to FH-bisimulation~\cite{fhbisim} we are interested  in finding refinement relations between configurations of two OA that contain variables and holes. When dealing with open systems it is common to define refinement in terms of a simulation relation.
We rely on a classical notion of simulation inspired from~\cite{fhbisim}. The idea is that two configurations related by a relation; if one state can do a transition, then the other can do a transition too. 
Like for bisimulation, refinement relation characterizes when two states are related, and this  characterisation is expressed as a predicate on the variables of the two automata.
Refinement defines conditions on a predicate $R$ such that $R\mpar{s_1, s_2}$ is a predicate (possibly involving variables of the automata) that is true when the state $s_1$ of $A_1$ refines the state $s_2$ of $A_2$.

@@

\begin{example} \label{example:fifo-system-one}
To illustrate the refinement simulation of OA, we consider a variation on the \texttt{prod-cons} example. Namely,  we suppose that  the two processes  \texttt{P} and  \texttt{Q} communicate through a one-place  buffer. Fig. \ref{Fig:SpecOA} shows the OA modelling this simpler version of the system, that we refer to as \texttt{Simprod-cons}. 
\begin{figure}[!t]
 \centering
   \includegraphics[width=.90\textwidth]{Figures/FIFOpen2.pdf}
   \caption{  The \texttt{Simprod-cons} Open automaton: the system using one-place  buffer.  \label{Fig:SpecOA}}
\end{figure}
We can easily check that this automaton simulates the one of Fig.~\ref{Fig:RefineOA}.
Indeed, one can see that  $R=\{(r_0,s_0) \mapsto \texttt{l}=\texttt{f}, (r_0,s_1) \mapsto \texttt{f}=\texttt{l}+1\% \texttt{N} \}$ is  a candidate simulation 
relation. It follows that  \texttt{prod-cons} $\leq$ \texttt{simprod-cons}. \qed
\end{example}

@@

%\begin{definition}[Non-blocking composition] \label{Def:Non-block}
%Let \(A_i = \OAg[i]\) with \( 0 \leq i \leq n\) be a family of open automata.
%Let \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) and $A= \OAg$.
%
%The composition \(A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) is non-blocking if   \(A \) has a reachability predicate such that, for each reachable configuration, if there is a possible transition in \(A_0\) then there is a possible transition in \(A\):
%\begin{multline*}
% \forall s = \mpar{s_0, s_1,..,s_n} \in S, V \uplus \biguplus_\subbox{t_0 \in \fOT{s_0}} \fvars{t_0} \vdash \\ \left(\reach{A}\mpar{s} \wedge \bigvee_\subbox{t_0 \in \fOT{s_0}} \fguard{t_0} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \right)
%\end{multline*}
%\end{definition}
\begin{definition}[Non-blocking composition] \label{Def:Non-block}
Let \(A_1 \defobject \OAg[1]\) and \( A_2 \defobject \OAg[2]\) be two OA.
Let $A= \OAg$ be the OA resulting from the composition \(A = A_1\subst{A_2}{k}\). 
The composition \(A_1\subst{A_2}{k}\) is non-blocking if   \(A \) has a reachability predicate such that, for each reachable configuration, if there is a possible transition in \(A_1\) then there is a possible transition in \(A\):
\begin{multline*}
 \forall s = \mpar{s_1, s_2} \in S, V \uplus \biguplus_\subbox{t \in \fOT{s_1}} \fvars{t} \vdash  \left(\reach{A}\mpar{s} \wedge \bigvee_\subbox{t \in \fOT{s_1}} \fguard{t} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \right)
\end{multline*}
\end{definition}
Like in the definition of refinement (Definition~\ref{def:HoleEqualSim}) we use guards to ensure that the transition can occur. 
In general, one would not want to only consider non-blocking composition as it may reveal a bit restrictive, but it is the best necessary condition that can be identified for compositionality of refinement. It will be used to prove composition theorems in  Section~\ref{sec:prop}. 
Proving refinement relation in absence of non-blocking composition can always be done specifically for the composed automaton under study.
%It is not sufficient to ensure the existence of equivalent transitions in general, but it will be sufficient in the context of refinement.

@@

\begin{property}[Relating simulations]\label{lem-sim-sim}
Hole-equal simulation is a particular case of hole-tracking refinement when  $J_1=J_2=H$.
\end{property}
\begin{example} Consider the automata of examples \ref{Example:fifo-system} and \ref{example:fifo-system-one}. As we saw above, \texttt{prod-cons} $\leq$ \texttt{simprod-cons}, therefore \texttt{prod-cons} $\leq_{\{\texttt{P,Q}\}}$ \texttt{simprod-cons}.
\end{example}
\TODO{potential ex 4: check if the example with a hole filled is a simulation of the example with holes? at least state \texttt{prod-cons} $\leq_{\{\texttt{P,Q}\}}$ \texttt{simprod-cons}}

@@

\begin{theorem}[Context refinement]\label{thm:ContextRefinement}
Let $A_1$, $A_2$ and $A_3$ be three OA with $\wrel{A_1}{A_2}{H}$. 
Let $J_3$ be the set of holes of $A_3$ and suppose that $k \in H$.
Suppose additionally that  \(A_1\subst{A_3}{k}\) is non-blocking.
We have: \[\wrel{A_1\subst{A_3}{k}}{A_2\subst{A_3}{k}}{J_3 \uplus H \setminus \mbrc{k}}\]
\end{theorem}

@@

\RAB{Pourquoi on parle bisimulation??}
The bisimulation relation that we consider is the one that makes $A_1$ and $A_2$ bisimilar, complemented with identity of configurations for $A_3$.
The proof relies on the fact that, by construction, all transitions of the composed automaton $A_1\subst{A_3}{k}$ are specified by open transitions of $A_1$. For the transitions that do not involve hole $k$, the transition of $A_1\subst{A_3}{k}$ is the same and simulation between $A_1$ and $A_2$ allows us to conclude directly. If the hole $k$ is involved the considered relation  implies that valuations in $A_3$ are equal (i.e., the value for each variable are the same in both valuations), after a transition we should obtain ``equal'' valuations because post-conditions are deterministic.

\begin{example}  Consider again the \texttt{prod-cons} and \texttt{Simprod-cons}  automata given in the examples above.  Since  \texttt{prod-cons} $\leq_{\{\texttt{P,Q}\}}$ \texttt{simprod-cons},  then according to Theorem \ref{thm:ContextRefinement}, 
\texttt{prod-cons[producer/P]} $\leq_{\{\texttt{Q}\}}$  \texttt{Simprod-cons[producer/P]}. 
The automaton of \texttt{prod-cons[producer/P]} is shown in Fig.~\ref{Fig:ComposeOA}.
The automaton resulting  from the composition of  \texttt{Simprod} and \texttt{producer} is bigger and not shown here. \qed
\end{example}

\begin{theorem}[Congruence]\label{thm:CongurenceRefinement}
Let $A_1$, $A_2$ and $A_3$ be three OA with $\wrel{A_2}{A_3}{H}$. 
  Let $J_1$ be the set of holes of $A_1$ and suppose that \(k \in J_1\). Suppose additionally that the composition $A_1\subst{A_2}{k}$ is non-blocking.
We have: \[\wrel{A_1\subst{A_2}{k}}{A_1\subst{A_3}{k}}{J_1 \uplus H \setminus \mbrc{k}}\]
\end{theorem}

Consequently, as the simulation is transitive \TODO{TODO Thm 4}
We can state the following theorem: 
\begin{theorem}[Composability]\label{thm:Composability}
Let $A_1$, $A_2$, $A_3$ and $A_4$ be four OA with $\wrel{A_1}{A_2}{H}$ and $\wrel{A_3}{A_4}{H'}$.   Let $J_1$ and $J_2$ be the set of holes of $A_1$ and $A_2$ resp. and suppose that \(k \in J_1 \cap J_2\). We have: \[\wrel{A_1\subst{A_3}{k}}{A_2\subst{A_4}{k}}{J_1 \uplus J_2 \uplus J_3 \uplus J_4 \uplus H \uplus H'\setminus \mbrc{k}}\]
\end{theorem}

\begin{example}
As an example of the use of this theorem, if we design a refined version of the producer process of Example~\ref{Example:prodandcons} called $\texttt{producer}^{\flat}$. According to  Theorem \ref{thm:Composability}, we have 
\texttt{prod-cons[producer/P]} $\leq_{\{\texttt{Q}\}}$  \texttt{Simprod-cons[$\texttt{producer}^{\flat}$/P]}. 
%as well as \texttt{prod-cons[P/producer]} $\leq_{\{\texttt{Q}\}}$  \texttt{Simprod-cons[P/$\texttt{producer}^{\flat}$]}.
\end{example}

@@

The simulation  ensures the preservation of   
safety properties as deadlock-freeness  and, more generally, all 
all linear temporal logic properties.  The issue is about the properties of simulation, in particular closure properties such as preservation under composition. Some works   have addressed the problem of non-introduction of blocking by composition. For example, in \cite{DIHEGO2020110598} it is presented a refinement relation on process algebra based on failure semantics that is basically different from the (bi)simulation \cite{10.5555/640428.640430}.

@@

\section{Proof of Transitivity for Refinement}\label{sec:proof-transitivity}
\begin{quote}
If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
\end{quote}

\proof 
If $\wrel{A_1}{A_2}{H}$ then there is $R_{12}$ a relation between states
of $A_1$ and of $A_2$;  If $\wrel{A_2}{A_3}{H'}$ then there is $R_{23}$ a relation between states of $A_2$ and of $A_3$. We build a relation between
 states of $A_1$ and of $A_3$ as follows:  for each pair of states $s_1$, $s_3$, for each state $s_2$ such that $R_{12}$ relates $s_1$ and $s_2$, and $R_{23}$ relates $s_2$ and $s_3$.
Let $R_{13}$ be the relation:\\
  \[R_{13}(s_1,s_3)=\bigvee_{s_2\in S_2}\mpar{R_{12}(s_1,s_2) \land R_{23}(s_2,s_3) } \]

@@

This is valid for all $s_{2p} \in (s_{2p})^{p\in P}$, then we have:
\begin{multline*}
 V_1 \uplus V_2 \uplus V_3  \vdash \underbrace{\bigvee_{{p\in P}}\mpar{R_{12}\mpar{s_1, s_{2p}} \wedge R_{23}\mpar{s_{2p},s_3}}}_{R_{13}(s_1,s_3)} \implies  ( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} )
 \lor¢ \lnot( \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} )
\end{multline*}
By removing the $A_2$ variables that have not effect on $A_2$ et $A_3$,
we get the desired result:
\begin{multline*}
 V_1 \uplus  V_3  \vdash R_{13}(s_1,s_3) \implies  ( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} )
 \lor¢ \lnot( \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} )
\end{multline*}
\qed
\end{enumerate}
\section{Proof of Context Refinement (Theorem~\ref{thm:ContextRefinement})}

\begin{quote}
Suppose that $\wrel{A_1}{A_2}{H}$, \(k \in H\) and that \(A_1\subst{A_3}{k}\) is non-blocking. We have: 
\[\wrel{A_1\subst{A_3}{k}}{A_2\subst{A_3}{k}}{J_3 \uplus H \setminus \mbrc{k}}\]
\end{quote}
\smallskip
Pictorially, the theorem states the following result:

@@

\section{Proof of Congruence}
\begin{quote}
Suppose $\wrel{A_2}{A_3}{H}$ and  $k \in H$.  and the composition $A_1\subst{A_2}{k}$ is non-blocking. We have: \[\wrel{A_1\subst{A_2}{k}}{A_1\subst{A_3}{k}}{J_1 \uplus H \setminus \mbrc{k}}\]
\end{quote}
\smallskip
Pictorially, the theorem states the following result:

\begin{center}
\includegraphics[scale=0.7]{Figures/Thm2intuition}
\end{center}
\proof
Let us denote by $A_{12}$ (resp. $A_{13}$) the OA resulting from $A_1\subst{A_2}{k}$ (resp. $A_1\subst{A_3}{k}$),  to prove the theorem it is sufficient to prove that there exists a relation between states of the two OA that satisfies the conditions of the Definition \ref{Def:OA-Refinement}.
